Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-05-03 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Still working on Chapter 5

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure  is a schematic of the cone calorimeter.

@@

The radiant flooring panel test method is known as ASTM~E648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}. A 1~m long test specimen is mounted horizontally beneath a pre-mixed air-gas-fueled radiant panel. The panel is inclined at 30~$^{\circ}$, and under standard operating conditions provide a heat flux on the test specimen varying from 10~kW/m$^2$ at the measurement point, centered 100~mm from the edge of the sample closest to the panel to 1~kW/m$^2$ at the farthest measurement point 900~mm from the edge of the sample closest to the panel. For these experiments a higher heat flux profile was required, so the gas air increased to provide an incident flux of approximately 21~kW/m$^2$ to 2~~kW/m$^2$. Figure~\ref{FRP_HF_Calibration}

A pilot flame from a mm long line burner is used to ignite the test sample near the edge of the sample closet to the panel. The objective of the test is to determine the critical radiant flux for a given sample.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/FRP_HF_Calibration} \\
	\caption[Incident heat flux profile in the radiant flooring panel test apparatus]{Incident heat flux profile along the centerline of the sample in the radiant flooring panel test apparatus}
	\label{FRP_HF_Calibration}
\end{figure}

@@

The vertical radiant panel is used as documented in ASTM~E162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that is 305~mm × 457~mm.

However for the purposes of this study a modified version of the test apparatus was used.  The radiant panel and and the radiant panel controls remained in place as documented in ASTM~E162. However the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux.  This trolley was used to hold a sample of painted gypsum wallboard were exposed to the radiant .   Figure

In the modified test configuration the exposed surface of the sample was parallel to the radiant panel. The radiant panel was calibrated and mapped with a heat flux gauge. With the flux gauge data, positions away from the radiant panel were identified to enable the samples to be exposed to a range of heat fluxes.  After a timed exposure the samples were examined and area of thermal damage to the painted paper surface of the gypsum board was compared with heat flux mapping.  This was another means of potentially obtaining a measure of the critical heat flux. Figure 2.3.11 illustrates the radiant panel test apparatus.

@@

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/RP_GB_Front} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  ]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  Each row of samples represents three replicate experiments at a given heat flux for 600s. Begining with the top row and moving down the exposure heat fluxes were 10, 12.5, 15, 17.5, 20, 22.5 and 25~kW/m$^2$.  }
	\label{RP_GB_Front}
\end{figure}

@@

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/RPGB_SET} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  ]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  Each row of samples represents three replicate experiments at a given heat flux for 600s. Begining with the top row and moving down the exposure heat fluxes were 10, 12.5, 15, 17.5, 20, 22.5 and 25~kW/m$^2$.  }
	\label{RPGB_SET}
\end{figure}

>>>>

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Cone_Cal} \\
	\caption[Sketch of the cone calorimeter]{Sketch of the cone calorimeter}
	\label{Cone_Cal}
\end{figure}

@@

The radiant flooring panel test method is known as ASTM~E648, Standard Method of Test for Critical Radiant Flux of Floor-Covering Systems Using a Radiant Heat Energy Source~\cite{ASTM_E648}.  The radiant flooring panel is used to measure the flame spread of a material while exposed to an incident radiant heat flux. The source of the heat flux is a pre-mixed air-gas-fueled radiant panel.  The radiating surface measures 305~mm by 457~mm. The radiating surface is inclined at 30~$^{\circ}$ above the horizontal. The test specimen is mounted horizontally beneath the radiant panel.  The surface of the test sample exposed to the radiant panel is 1~m long and 0.2~m wide. Under standard operating conditions the radiant panel provides a heat flux distribution on the test specimen varying from 10~kW/m$^2$ at the measurement point, centered 100~mm from the edge of the sample closest to the panel to 1~kW/m$^2$ at the farthest measurement point 900~mm from the edge of the sample closest to the panel. A schematic of the radiant flooring panel apparatus is shown in Figure~\ref{FLR_Radpan}.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/FLR_Radpan} \\
	\caption[Schematic of the radiant flooring panel apparatus]{Schematic of the radiant flooring panel apparatus}
	\label{FLR_Radpan}
\end{figure}

 For these experiments a higher heat flux profile was required to propagate flame spread on the painted gypsum wallboard samples, so the air-gas flow supplying the radiant panel was increased to provide an incident flux ranging from approximately 21~kW/m$^2$ at the 100~mm position to 2~kW/m$^2$ at the 900~mm position.  The incident heat flux curve used in these experiments is shown in Figure~\ref{FRP_HF_Calibration}.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/FRP_HF_Calibration} \\
	\caption[Incident heat flux profile in the radiant flooring panel test apparatus]{Incident heat flux profile along the centerline of the sample in the radiant flooring panel test apparatus}
	\label{FRP_HF_Calibration}
\end{figure}

A pilot flame from a 254~mm long wide stainless steel burner tube, as specified in ASTM~648, was used to ignite the test sample near the edge of the sample closet to the panel~\cite{ASTM_E648}. The objective of these experiments was to determine the critical radiant flux for the painted gypsum wallboard samples.  The critical radiant flux corresponds to the flux at the point of maximum flame propagation, or the point on the sample where the flame spread stopped. For this study the samples examined were cut from painted gypsum wallboard panels that were prepared from the supply of panels that were used in the compartment experiments.

@@

The vertical radiant panel used is documented in ASTM~E162 and normally provides a measure of downward, opposed-flow flame spread~\cite{ASTM_E162}.  The apparatus measures the surface flammability of materials using a pre-mixed air-gas-fired radiant panel that has a radiating surface 305~mm wide by 457~mm high.

For the purposes of this study a modified version of the test apparatus was used.  The radiant panel and the radiant panel controls remained in place as documented in ASTM~E162. However the sample mounting system was changed. A rail and trolley system was developed by Lawson and Twilley to expose fire fighter protective equipment to various heat fluxes by being able to move the test specimen to a predetermined position that correlated with a given heat flux.  This trolley was used to hold a sample of painted gypsum wallboard that was 362~mm square.  The samples were vertical and parallel to the radiant panel. Figure~\ref{Rad_Panel} has photographs which show the radiant panel test arrangement.

\begin{figure}
	\centering
	\includegraphics[width=2.92in]{../Figures/RAD_pan}
	\includegraphics[width=2.08in]{../Figures/RAD_pan_iso} \\
	\caption[Photographs of the radiant panel experimental arrangement]{Photographs of the radiant panel experimental arrangement, on the left is a photograph which shows the radiant panel on the right, the aluminum heat shield in the center and the moveable mounting frame for the gypsum wallboard sample on the left.  The photograph on the right shows an isometric view of the test arrangement.}
	\label{Rad_Panel}
\end{figure}

Heat flux gauge measurements, taken at the center position of the sample, were used to determine positions on the trolley track for a specific radiant exposure. Positions were marked for 10, 12.5, 15, 17.5, 20, 22.5, and 25~kW/m$^2$. The area surrounding the center of the maintained the desired heat flux within at least 70~mm on either side of the center point and 100~mm above and below the center point.  In other words, the area around the center point of the sample had heat flux measures similar, within the uncertainty of the device, to the center point measurement. Outside of the central area of the samples, the heat flux would decrease as the position approached the edge of the sample.  The center point heat flux measures vs the distance from the radiating surface is provided in Figure~\ref{Rad_Panel_Cal}

To conduct the experiments, the radiant panel was ignitied and allowed to pre-heat for at least 60 minutes.  The water cooled, Schmidt-Boelter total heat flux gauge was mounted in the center the sample holder, and the heat flux positions were confirmed.  Just before the sample was mounted on the movable trolley, an aluminium heat shield was placed between the radiant panel and the sample holder.  Once the sample was mounted and the sample was moved to the appropriate distance from the radiant panel based on the heat flux desired for the experiment, the heat shield was removed that the experiment began.  Each sample was exposed for 600 seconds. Each sample had a single bare-bead, Chromel-Alumel (type K) thermocouple, with a 0.5~mm nominal diameter bead in contact with the center point of each sample. During the experiments the samples were monitored for temperature and ignition.    examined and area of thermal damage to the painted paper surface of the gypsum board was compared with heat flux mapping.  This was another means of potentially obtaining a measure of the critical heat flux. .11

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/Rad_Panel_Cal} \\
	\caption[Incident heat flux profile based on the distance between the sample face and the radiant panel]{Incident heat flux profile based on the distance between the sample face and the radiant panel}
	\label{Rad_Panel_Cal}
\end{figure}

@@

Exposing the painted gypsum wallboard samples to the vertical radiant panel provided a wide range physical damage to samples as shown in Figure~\ref{RPGB_SET}.  However there based on visual observations there were no flaming ignitions.  One of the samples at 15~kW/m$^2$, small sections of the paint/paper fuel would peel or partially separate from the gypsum core and would begin to glow.  This process became more prevalent as the heat fluxes increased. There are no photos of the samples that were exposed to 25~kW/m$^2$.  The radiant exposure caused the samples to collapse as they were being removed from the holder.  Figure~\ref{RP_GB_Front} is a graph of the averaged time vs temperature curves for the three replicates at the 7 different heat flux levels.  Notice that the samples exposure at 17.5~kW/m$^2$ and higher have a temperature increase to a peak temperature followed by a decrease.  This trend may be may indicative of combustion of the paint and paper adding to the energy at on the face of the sample.

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/RPGB_SET} \\
	\caption[Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  ]{Photographs of the fire effects on the instrumented gypsum wallboard exposed to radiant heat from a natural gas fired panel.  Each row of samples represents three replicate experiments at a given heat flux. Beginning with the top row and moving down the exposure heat fluxes were 10, 12.5, 15, 17.5, 20, and 22.5~kW/m$^2$.}
	\label{RPGB_SET}
\end{figure}

\begin{figure}[p]
	\centering
	\includegraphics[width=\textwidth]{../Figures/RP_GB_Front} \\
	\caption[Time vs. temperature curves for the gypsum wallboard samples exposed to radiant heat from a natural gas fired panel]{Time vs. temperature curves for the gypsum wallboard samples exposed to radiant heat from a natural gas fired panel at heat fluxes of 10, 12.5, 15, 17.5, 20, 22.5 and 25~kW/m$^2$.}
	\label{RP_GB_Front}
\end{figure}

@@

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 9~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and sample nearing the flameout position are provided in Figure~\ref{FRP_Test_Photos}.

\begin{figure}
	\centering
	\includegraphics[width=1.8in]{../Figures/FRP_Test_Photo1}
	\includegraphics[width=3.2in]{../Figures/FRP_Test_Photo2} \\
	\caption[Photographs of the radiant flooring panel experiments]{Photographs of the radiant flooring panel experiments, on the left is a photograph aimed at the radiant panel showing the pilot flame on the sample as well as the flame front on the sample.  The photograph on the right show the flame front on the sample just prior to flame out.}
	\label{FRP_Test_Photos}
\end{figure}
